Tackling discrepancies in freshwater invertebrate trait databases and taxonomic resolution: Harmonising across continents and trait aggregation

### 2021-01-26 Paper/Draft_harmonization_paper.tex

1. Use of invertebrate traits rather than species composition is expected to facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere. % OLD: because of their consistent response to environmental gradients
In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, for freshwater invertebrate traits standardised definitions are lacking, impeding comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions between databases from different databases might influence trait-environment relationships. Another issue is that if the taxonomic resolution of observational taxonomic data is less precise than in trait databases, the traits in databases require aggregation before they can be linked to the observational data.
% Alternative from Co-Authors: Another issue is that we do not know how taxonomic resolution affects the characterization of traits and hence the comparability of different datasets.
% Another issue is that if aggregation of traits is necessary, i.e. when observational taxonomic data is less precise than available trait data, no guidance exists which method under which circumstances can

@@

3. Between the trait databases, grouping features differed in terms of their classification into traits and their scope leading to discrepancies how traits are defined. Also, the coding to describe traits varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, especially approaches using the median to aggregate from species to family-level. Grouping feature harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects did occur in some cases because of harmonisation.

@@

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits are measurable properties of an organism (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{heino_jani_macroecological_2013}). Traits are assumed to be adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and represent direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}).
Besides providing more mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analysis because the variability in trait responses is lower than in taxonomic responses (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Traits of individual freshwater invertebrate are difficult to determine because - unlike plant traits - they are often difficult to measure directly. For example, to describe feeding habits requires that we understand both mouthpart morphology and the specific food that is consumed, and the organisms function within its community (\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors and as trait metrics in biomonitoring (\cite{bhowmik_large_2015, menezes_beyond_2010, poff_developing_2010, szocs_effects_2014}).

In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases for various continents (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents arguably enables comparisons of trait variation and testing if variation in trait structure is consistently related to environmental factors across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterized trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different continents are rare.

@@

In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, where a grouping feature is defined as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species". Thus, we use the term grouping feature in place of which many studies use the term "trait", and the term trait instead of "trait state", "modality" or "trait category". 
% Ben suggest here: For  example  e.g., predator, shredder, etc. Are traits of the grouping feature ..... and removing predator, shredder, etc. from the sentence above.
Traits can be described using different codings (binary, fuzzy) that represent the prevalence of a characteristic in an organism.

@@

These challenges when working with or synthesizing trait data motivated us to (1) investigate discrepancies in trait definition between trait databases from Europe, North America, Australia, and New Zealand. As a basis for further evaluating different trait aggregation methods and how harmonising definition discrepancies between grouping features can influence trait-environment relationships, we developed 4 novel invertebrate trait datasets from these existing trait databases by harmonising 7 grouping features. (2) We then evaluated different trait aggregation methods by comparing aggregated traits with traits assigned by experts at family-level for two of the established trait datasets. (3) Finally, we re-analysed data from a study of salinisation effects on biological traits (\cite{szocs_effects_2014}) using data from the newly established harmonised European trait dataset and aggregated traits. By comparing the re-analysis's findings with those of \citet{szocs_effects_2014}, we investigated how harmonising and aggregating trait data can modify the outcome of trait-environment relationships.

>>>>

1. Use of invertebrate traits rather than species composition is expected to facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere. % OLD: because of their consistent response to environmental gradients
In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, for freshwater invertebrate traits standardised definitions are lacking, impeding comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions between databases might influence trait-environment relationships. Another issue is that if aggregation of traits is necessary, i.e. when observational taxonomic data is less precise than available trait data, no guidance exists which method under which circumstances can 
% Alternative from Co-Authors: Another issue is that we do not know how taxonomic resolution affects the characterization of traits and hence the comparability of different datasets.

@@

3. Between the trait databases, grouping features differed in terms of their classification into traits and their focus leading to discrepancies how traits are defined. Also, the coding to describe traits varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, especially approaches using the median to aggregate from species to family-level. Harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects did occur in some cases because of harmonisation.

@@

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits are measurable properties of an organism (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{heino_jani_macroecological_2013}). Traits are assumed to be adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and represent direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}).
Besides providing more mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analysis because trait responses are less constrained by biogeographic boundaries and taxon distributional areas (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Traits of individual freshwater invertebrate are difficult to determine because - unlike plant traits - they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behavior, and the specific food that is consumed(\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}).

In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases for various continents (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents arguably enables comparisons of trait variation and testing if variation in trait structure is consistently related to environmental factors across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterized trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesize invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

@@

In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, where a grouping feature is defined as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species". Thus, we use the term grouping feature in place of which many studies use the term "trait", and the term trait instead of "trait state", "modality" or "trait category". 
% Ben suggest here: For  example  e.g., predator, shredder, etc. Are traits of the grouping feature ..... and removing predator, shredder, etc. from the sentence above.
Traits can be described using different codings (binary, fuzzy) that represent the expression of a characteristic in an organism.

@@

These challenges when working with or synthesizing trait data motivated us to (1) investi<gate discrepancies in trait definition between trait databases from Europe, North America, Australia, and New Zealand. As a basis for further evaluating different trait aggregation methods and how harmonising definition discrepancies between grouping features can influence trait-environment relationships, we developed 4 novel invertebrate trait datasets from these existing trait databases by harmonising 7 grouping features. (2) We then evaluated different trait aggregation methods by comparing aggregated traits with traits assigned by experts at family-level for two of the established trait datasets. (3) Finally, we re-analysed data from a study of salinisation effects on biological traits (\cite{szocs_effects_2014}) using data from the newly established harmonised European trait dataset and aggregated traits. By comparing the re-analysis's findings with those of \citet{szocs_effects_2014}, we investigated how harmonising and aggregating trait data can modify the outcome of trait-environment relationships.
